---
title: "Data wrangling for online retail shop"
output: html_notebook
---

# loading readxl library so that the program can read the dataset which is in xlsx extension 
```{r}
library(readxl)
```
```{r}
library(dplyr)
```
# for data visualization while cleaning the data
```{r}
library(ggplot2)
```




# since the dataset is in .xlsx extension so loading the data set in xlsx form
```{r}
data_file <- read_xlsx("online_retail_II.xlsx")
```
# saving the dataset in csv format so that data manipulation can be done to the data.
```{r}
write.csv(data_file, "online_retail.csv", row.names = FALSE)
```
# reading the saved data set and storing it in the data_file variable
```{r}
data_file <- read.csv("online_retail.csv")
```
# heading into to the data and looking at the frist few data
```{r}
head(data_file)
```
# Displaying the basic information about the dataset
```{r}
str(data_file)
```

```{r}
# Checking for missing values in the data frame
missing_value <- is.na(data_file)

# Locating the position of the missing value
missing_position <- which(missing_value, arr.ind = TRUE)

rows_with_missing_value <- unique(missing_position[,1])
print(data_file[rows_with_missing_value, ])
```

```{r}
# Total rows of the dataset
paste("total rows of the dataset: ", nrow(data_file))

# total no. of missing values in the dataset
paste("No of rows with missing value: ", sum(is.na(data_file)))
```
# removing the null values from the dataset and storing the non-null valued dataset in the new_data_file variable (data frame).
```{r}
new_data_file <- na.omit(data_file)
```

```{r}
paste("total dataset with non-missing value: ",nrow(new_data_file ))
paste("No of missing data after omiting the nan variables: ",
sum(is.na(new_data_file)))
```
# looking at the summary of the logical object created in the new_data_set, and taking the summary of the null/nan variables.
```{r}
summary(is.na(new_data_file))
```
```{r}
# Checking if the dataset contains any duplicte vectors/values
paste("total no. of rows in the dataset: ",nrow(new_data_file ))
paste("No of duplicate values in the dataset: ", sum(duplicated(new_data_file)))
```
```{r}
# removing duplicate rows
new_data_file <- new_data_file %>% distinct()
```

```{r}
paste("total no. of rows in the dataset: ",nrow(new_data_file ))
paste("No of duplicate values in the dataset: ", sum(duplicated(new_data_file)))
```
# taking a look at the matrices of the new_data_file, to determine the data type of the columns in the dataset.
```{r}
matrix(new_data_file)
```

```{r}
# Parsing the InvoiceDate column into a proper date-time format
data_file$InvoiceDate <- as.POSIXct(data_file$InvoiceDate, format="%Y-%m-%d %H:%M:%S")
```

```{r}
# Removing leading/trailing whitespace from product descriptions
data_file_clean$Description <- trimws(data_file_clean$Description)

# Converting all descriptions to lowercase for consistency
data_file_clean$Description <- tolower(data_file_clean$Description)

```

```{r}
# Adding TotalPrice as a new column (Quantity * Price)
data_file$TotalPrice <- data_file$Quantity * data_file$Price
```

```{r}
# Z-score calculation for Price to detect outliers
data_file$z_score <- scale(data_file$Price)
```

```{r}
# Visualizing the Price distribution before outlier handling
ggplot(data_file, aes(x = factor(0), y = Price)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(title = "Box Plot of Prices Before Outlier Handling", x = "Price", y = "Value")
```

```{r}
# Detecting outliers based on z-scores > 3 or < -3
outliers <- data_file %>% filter(abs(z_score) > 3)
print(paste("Number of outliers detected: ", nrow(outliers)))
```
```{r}
# Removing outliers
data_file_clean <- data_file %>% filter(abs(z_score) <= 3)
print(paste("Total rows after removing outliers: ", nrow(data_file_clean)))

```
```{r}
# Visualizing the Price distribution after outlier removal
ggplot(data_file_clean, aes(x = factor(0), y = Price)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(title = "Box Plot of Prices After Outlier Removal", x = "Price", y = "Value")

```
# Since, Negative values in Quantity or Price might represent returns, refunds, or errors. Filtering the negative values for Quantity and Price as we are not analyzing return or cancellation for this data analysis.
# more focused on sales analysis.
```{r}
# Checking for negative quantities (indicating returns or potential errors)
negative_quantities <- data_file_clean %>% filter(Quantity < 0)
print(paste("Number of transactions with negative quantities: ", nrow(negative_quantities)))

# Removing the negative quantities
data_file_clean <- data_file_clean %>% filter(Quantity >= 0)
```

```{r}
# Recalculating the TotalPrice to ensure it's consistent
data_file_clean$TotalPrice <- data_file_clean$Quantity * data_file_clean$Price
```
```{r}
# Z-score calculation for Quantity and TotalPrice
data_file_clean$z_score_quantity <- scale(data_file_clean$Quantity)
data_file_clean$z_score_total_price <- scale(data_file_clean$TotalPrice)
```
```{r}
# Visualizing outliers in Quantity
ggplot(data_file_clean, aes(x = factor(0), y = Quantity)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(title = "Box Plot of Quantity (Before Outlier Handling)", x = "Quantity", y = "Value")
```
```{r}
# Visualizing outliers in TotalPrice
ggplot(data_file_clean, aes(x = factor(0), y = TotalPrice)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(title = "Box Plot of TotalPrice (Before Outlier Handling)", x = "TotalPrice", y = "Value")

```

```{r}
# Detecting outliers using z-scores > 3 or < -3
quantity_outliers <- data_file_clean %>% filter(abs(z_score_quantity) > 3)
total_price_outliers <- data_file_clean %>% filter(abs(z_score_total_price) > 3)

```
```{r}
# Display number of outliers detected
print(paste("Number of Quantity outliers detected: ", nrow(quantity_outliers)))
print(paste("Number of TotalPrice outliers detected: ", nrow(total_price_outliers)))

```
```{r}
#Capping outliers at the 95th percentile

# Capping for Quantity
quantity_cap <- quantile(data_file_clean$Quantity, 0.95)
data_file_clean$Quantity_capped <- ifelse(data_file_clean$Quantity > quantity_cap, quantity_cap, data_file_clean$Quantity)

# Capping for TotalPrice
total_price_cap <- quantile(data_file_clean$TotalPrice, 0.95)
data_file_clean$TotalPrice_capped <- ifelse(data_file_clean$TotalPrice > total_price_cap, total_price_cap, data_file_clean$TotalPrice)

```
```{r}
# Boxplot for Quantity after outlier handling
ggplot(data_file_clean, aes(x = factor(0), y = Quantity)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(title = "Box Plot of Quantity After Outlier Handling", x = "Quantity", y = "Value")

```
```{r}
# Boxplot for TotalPrice after outlier handling
ggplot(data_file_clean, aes(x = factor(0), y = TotalPrice)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(title = "Box Plot of TotalPrice After Outlier Handling", x = "TotalPrice", y = "Value")

```



```{r}
# Rechecking the structure and summary of the cleaned dataset
summary(data_file_clean)
```
```{r}
# Final output after handling outliers
write.csv(data_file_clean, "cleaned_online_retail.csv", row.names = FALSE)
```


### making the data fit for data visualization ###

